---
title: "full_demo"
author: "Jinlu Liu"
date: '2024-02-29'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

## A full demonstration of how to use the normHDP algorithm.

#### First step is to Specify width and length: width is the number of chains to run and length is the total number of iterations in each run. We first obtain posterior samples of the allocations (z). Below is an example with 100 chains and 100 iterations in each chain, 30 components and running with 4 cores.

```{r}

Depth <- 100
Width <- 100

Z_trace_all <- lapply(1:Width,
                      function(w){

                        normHDP_mcmc(Y = Y,
                                     J = 30,
                                     number_iter = Depth,
                                     thinning = 1,
                                     burn_in = 0,
                                     quadratic = TRUE,
                                     print_Z = TRUE,
                                     num.cores = 4,
                                     save.only.z = TRUE,
                                     run.on.pc = TRUE)

                      })

```

#### For some general analysis:

```{r}

# Function below computes the optimized clustering estimate
Z_opt <- consensus_clustering(Z_output = Z_trace_all,
                              num.cores = 4,
                              run.on.pc = TRUE)

# Clustering estimate
Z_opt$cluster.estimate

# Posterior similarity matrix: the first item shows the posterior
# similarity between cells within the one dataset, or across two datasets.
# The second item shows the posterior similarity between cells from all
# datasets
Z_opt$psm

# Heat map to show posterior similarity matrix
# The function outputs 2 heat-maps. The first heat-map considers
# cells from different datasets separately. The second heat-map
# considers cells from all datasets together. Also need to give name
# to each of the two plots. Ideally, for datasets with a large number of
# observations, we need to output the heat-map before viewing.
plotpsm(psm.ind = Z_opt$psm$psm.within,
        psm.tot = Z_opt$psm$psm.combined,
        method = 'complete',
        plot1.name = 'psm1',
        plot2.name = 'psm2',
        xlab = 'Cell',
        ylab = 'Cell')

# Plots to summarize the clustering estimate. In the case when the number of datasets
# equal to 2, the function outputs 2 plots: plot 1 to show the number of cells
# in each cluster, and plot 2 to show the proportion of cells from each dataset
# in each cluster. In the case when the number of dataset is not equal to two, only
# plot 1 will be shown
cluster_summary(Z = Z_opt$cluster.estimate,
                data.name = c('data1',
                              'data2'))


# Based on the estimated z, we can obtain samples of all other parameters.
# The code below uses 10 cores for parallel computing. Need to check the
# number of available cores before running the code below.
MCMC_run <- normHDP_mcmc_fixed_z(Y = Y,
                                 Z = Z_opt$cluster.estimate,
                                 number_iter = 5000,
                                 thinning = 5,
                                 burn_in = 2000,
                                 quadratic = TRUE,
                                 num.cores = 4,
                                 iter_update = 1,
                                 run.on.pc = TRUE)

# To observe the effect of changing chain width and length on the mean absolute
# difference of posterior similarity matrix
concensus_clustering_plot(Z_output = Z_trace_all,
                          W.breaks = c(10,30,50,70,90,100),
                          L.breaks = c(10,20,30,40,50,60,70,80,90,100),
                          num.cores = 10,
                          run.on.pc = TRUE)

# Compare posterior estimated capture efficiency with bayNorm estimated
# capture efficiency
beta_compare(Y = Y,
             normHDP_post_output = MCMC_run,
             data.names = c('data1','data2'))

# Difference in component probabilities between mouse in the case when there are
# only 2 datasets, clusters are grouped into 3 types; over-represented, under-represented
# and stable clusters
difference_in_component <- difference_in_p_jd2(p_jd_trace = MCMC_run$P_J_D_output)

# Distribution of posterior estimated capture efficiency for each cluster
beta_analysis(normHDP_post_output = MCMC_run,
              cluster.labels = difference_in_component$all_obs)


# Posterior relationship between mean expressions and dispersions on the log-scale
# for each cluster.
mu_and_phi_plot(normHDP_post_output = MCMC_run,
                row.n = 3)



```

#### To find global marker genes, we use the following method:

```{r}

gmg_df <- global_marker_genes(normHDP_post_output = MCMC_run,
                              threshold = list('mu' = 2.5,
                                               'phi' = 2.5),
                              num.cores = 4,
                              run.on.pc = TRUE)


# The function output a list of 4 items:
#
# 1) marker_DE: data frame to show the tail probability, absolute LFC and classification of
# global marker genes using estimated mean expressions.
# 2) marker_DD: data frame to show the tail probability, absolute LFC and classification of
# global marker genes using estimated dispersions.
# 3) alpha_M: threshold for the classification of global differentially expressed genes.
# 4) alpha_D: threshold for the classification of global differentially dispersed genes.


# Scatter plot to compare absolute LFC and tail probabilities. In addition,
# the function also outputs a bar chart to summarize the number of each
# type of the marker genes
global_marker_genes_plot(gmg_output = gmg_df)

# The function outputs 4 graphs. The first 2 graphs shows the estimated mean
# expression and dispersions reordered by the tail probabilities. The final
# 2 graphs shows the standardized estimated mean expressions and dispersions
# of global marker genes.
global_marker_genes_heatmap(gmg_output = gmg_df,
                            normHDP_post_output = MCMC_run)


```

#### To find local marker genes, we use the following method:
```{r}

lmg_df <- local_marker_genes(normHDP_post_output = MCMC_run,
                             threshold = list('mu' = 0.8,
                                              'phi' = 0.8),
                             num.cores = 4,
                             run.on.pc = TRUE)

# The function output a list of 4 items:
#
# 1) marker_DE: data frame to show the tail probability, absolute LFC and classification of
# local marker genes using estimated mean expressions.
# 2) marker_DD: data frame to show the tail probability, absolute LFC and classification of
# local marker genes using estimated dispersions.
# 3) alpha_M: threshold for the classification of local differentially expressed genes.
# 4) alpha_D: threshold for the classification of local differentially dispersed genes.


# Scatter plot to compare absolute LFC and tail probabilities in the
# local marker genes case. Bar chart to show the number of local marker
# genes in each cluster
local_marker_genes_plot(lmg_output = lmg_df)

# Heat map of estimated mean expressions and dispersions for the local
# marker genes in each cluster. Genes are reordered by the tail probabilities.
local_marker_genes_heatmap(lmg_output = lmg_df,
                           normHDP_post_output = MCMC_run)

```


#### To visualize estimated latent counts:
```{r}

latent_counts_Y <- latent_counts(Y = Y,
                                 normHDP_post_output = MCMC_run,
                                 gmg_output = gmg_df,
                                 num.cores = 4,
                                 run.on.pc = TRUE)

# In the case when the global marker genes detail is not given, the function output 6 items:
#
# 1) matrix.output: the estimated latent counts with reordered cells by the estimated clustering.
# 2) index.solid: index used to separate cells from different clusters when plotting.
# 3) index.dashed: index used to separate cells from different datasets when plotting.
# 4) Y_latent: estimated latent count without reordering of cells and genes.
# 5) Z: the optimal clustering
# 6) G: total number of genes
#
# In the case when the global marker genes detail is given, then item 1) from the above list is
# replaced by:
#
# 1) Y_latent_DE: the estimated latent counts with reordered cells and genes by tail probabilities
# of mean expressions.
# 2) Y_latent_DD: the estimated latent counts with reordered cells and genes by tail probabilities
# of dispersions.
# 3) DE_number: total number of global differentially expressed genes.
# 4) DD_number: total number of global differentially dispersed genes.

# Heat-map of case with global marker genes detail not given
latent_counts_plot(latent_counts_output = latent_counts_Y,
                   file.name.1 = 'latent_counts')

# Heat-map of case with global marker genes detail given
latent_counts_plot(latent_counts_output = latent_counts_Y,
                   file.name.1 = 'latent_counts_DE',
                   file.name.2 = 'latent_counts_DD')

```

#### To visualize observed counts:
```{r}

observed_counts_df <- observed_counts(normHDP_post_output = MCMC_run,
                                      Y = Y,
                                      gmg_output = gmg_df)


# In the case when the global marker genes detail is not given, the function output 4 items:
#
# 1) matrix.output: the estimated latent counts with reordered cells by the estimated clustering.
# 2) index.solid: index used to separate cells from different clusters when plotting.
# 3) index.dashed: index used to separate cells from different datasets when plotting.
# 4) Z: optimized allocation
#
# In the case when the global marker genes detail is given, then item 1) from the above list is
# replaced by:
#
# 1) Y_DE: observed counts with reordered cells and genes by tail probabilities
# of mean expressions.
# 2) Y_DD: observed counts with reordered cells and genes by tail probabilities
# of dispersions.
# 3) DE_number: total number of global differentially expressed genes.
# 4) DD_number: total number of global differentially dispersed genes.

# In the case when the global marker genes detail is given, the function outputs 2 heat-maps,
# hence need to give 2 names. In the case when the global marer genes detail is not given, then
# only file.name.1. needs to be given.
observed_counts_plot(observed_counts_output = observed_counts_df,
                     file.name.1 = 'DE_reorder.png',
                     file.name.2 = 'DD_reorder.png')

```


#### For some t-SNE plot:
```{r}

# t-SNE plot using all genes
tsne_plot(normHDP_post_output = MCMC_run,
          Y_latent = latent_counts_Y$Y_latent,
          Y = Y)

# t-SNE plot using global marker genes
tsne_global_marker_genes(gmg_output = gmg_df,
                         Y = Y,
                         Y_latent = latent_counts_Y$Y_latent,
                         normHDP_post_output = MCMC_run)

```


#### For posterior predictive checks:
```{r}


# Mixed posterior predictive check: simulate dispersion using all the dependent parameters under log-Normal.
# Standard posterior predictive check: using posterior draws of dispersions.

# Single replicated dataset - mixed posterior predictive check
ppc_single_plot(normHDP_post_output = MCMC_run,
                Y = Y,
                data.name = c('data1', 'data2'))


# Multiple replicated datasets - mixed posterior predictive check
ppc_multiple_df <- ppc_multiple(normHDP_post_output = MCMC_run,
                                Y = Y,
                                number_rep = 100,
                                num.cores = 4,
                                run.on.pc = TRUE)

ppc_multiple_plot(ppc_multiple_df = ppc_multiple_df,
                  title = c('data1', 'data2'))

# Multiple replicated dataset, calculate standard p-value of posterior predictive check
ppp_df_standard <- ppp_calculation(normHDP_post_output = MCMC_run,
                                   Y = Y,
                                   num.cores = 4,
                                   rep.number = 100,
                                   type = 'standard',
                                   run.on.pc = TRUE)

# Multiple replicated dataset, calculate mixed p-value of posterior predictive check
ppp_df_mixed <- ppp_calculation(normHDP_post_output = MCMC_run,
                                Y = Y,
                                num.cores = 4,
                                rep.number = 100,
                                type = 'mixed',
                                run.on.pc = TRUE)


# Plot of distribution of p-values
p_value_plot(ppc_output = ppp_df_standard)
p_value_plot(ppc_output = ppp_df_mixed)

```


#### To focus on visualizing the 'important' genes:
```{r}

heatmap_important_genes(important_genes = important_genes,
                        normHDP_post_output = MCMC_run,
                        gene_list = Y_genes)

```
